How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-19 2.25-formula.Rmd
Add files via upload
# updated references with DOI and discussion table

## Why use a formula to structure each chapter or paper? {#formula}

@@

### So what is the formula?

If we were in my office, at this point in the meeting I usually go to my white board and start drawing funnels (as a metaphor) and blocks. However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [@mcgill20165]. But you can also find other useful material in a paper by Sheela Turbek and colleagues [-@turbek2016scientific], and in a paper by Kevin Plaxco [-@plaxco2010art]. If you are of the Twitter persuasion, you get lots of useful tips and links from WriteThatPhD and Write4Research.

@@

#### My take on how to write a formula paper

@@

#### Start with an outline
__So now that you've got the idea that writing to a formula is a good idea, what next? __
Next, I suggest that you begin to plan out your formula as a bullet-point outline. Start by writing a general sentence for the subject of each paragraph. Then use sub-points to plan each sentence within the paragraph. Lastly, annotate these with citations that you want to use. Personally, I'd prefer to see the bare outline, and then a fleshed outline before you start the writing proper. This would allow the advisor and student to get on the same page to be sure that what is planned is thought out.

@@

#### Next flesh out the outline with citations

@@

#### Lastly it’s time to write the text

>>>>

# Why use a formula to structure each chapter or paper? {#formula}

@@

## So what is the formula?

If we were in my office, at this point in the meeting I usually go to my white board and start drawing funnels (as a metaphor) and blocks. However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [@mcgill20165]. But you can also find other useful material in other papers [-@turbek2016scientific; @plaxco2010art @tomaska2007teaching]. If you are of the Twitter persuasion, you get lots of useful tips and links from WriteThatPhD and Write4Research.

@@

### My take on how to write a formula paper

@@

### Start with an outline
__So now that you've got the idea that writing to a formula is a good idea, what next? __
Next, I suggest that you begin to plan out your formula as a bullet-point outline. Start by writing a general sentence for the subject of each paragraph. Then use sub-points to plan each sentence within the paragraph. Lastly, annotate these with citations that you want to use. Personally, I'd prefer to see the bare outline, and then a fleshed outline before you start the writing proper. This would allow the advisor and student to get on the same page to be sure that what is planned is thought out.

@@

### Next flesh out the outline with citations

@@

### Lastly it’s time to write the text

### 2021-04-19 3.06-Discussion.Rmd
Add files via upload
# updated references with DOI and discussion table

## The Discussion {#discussion}

@@

Again, I'd like to repeat that the convention described above is a formula that is intended to help you getting started with writing. It is not the only way to write, and if you feel confident that another way is better then I encourage you to explore this with the support of your advisor. Be imaginative and don't feel that you are constrained or compelled to do anything in a formulaic way.

### First paragraph of the discussion
Your discussion begins by you responding to your hypothesis, clearly stating whether or not it has been accepted, and putting this into the wider context of the study (i.e. paragraph one or two of the [introduction](introduction.html) with relevant literature). You can then follow these statements by emphasising what you consider to be the most important finding, and explain how it adds to existing literature. However, don’t be tempted to over-interpret your results, or claim that they mean more than they do ([see section on speculation](speculation.html) below).

@@

### To sub-section or not to sub-section the discussion?
My preference is to plan the discussion before you write it, just as you did for the [introduction](introduction.html) using a bullet pointed list. This will provide you with logical sub-section headers for the discussion under which you can write the first draft. When your chapter has a simple aim that is easily communicated, I’d suggest deleting these sub-section headings before you finish. However, many studies are more complex and contain multiple experiments or evidential approaches. It is then sometimes wise to leave sub-sections in your discussion so that your reader can more easily follow the text. Where possible, these should be the same sub-sections that you have broken your [methods](methods.html) and [results](results.html) into, especially where these relate to specific hypotheses or aims. Or it may be more appropriate to discuss the different approaches separately, specifically when the literature that you refer to falls into different groups.

@@

@@

### Caveats and limitations
An important aspect of the discussion is to consider how the interpretation of your [results](results.html) may be incorrect. For example, if you have done an experiment, how well controlled was it and how well could it be considered to scale up to real world interactions? Could you have measured other variables? Almost every study will have caveats and limitations, and it is very important that you report them in a considered approach. You can also add ideas on how to improve your approach to eliminate important caveats that you have identified, even if this means suggesting something that adds evidence from a completely different field of biology.

@@

### Should you speculate in the discussion? {#speculation}
Reviewers will often be unhappy with speculation in the discussion section. Speculation isn’t that hard to spot, as it occurs when you make claims for which your [results](results.html) have no foundation. I think that it is healthy to have one or two statements that are speculative, but clearly labelled as such. After all, after writing this paper, you are going to be one of the world experts in the topic, and thus your deeper understanding is often worth relating to the reader. However, I suggest that you speculate in combination with suggesting what work could be done in future. If you really feel that the point has to be made, you must clearly label it as speculative.

@@

### Don’t beat up on others
Your [results](results.html) may show that other researchers were wrong with their interpretation or findings. Whatever you may think of them, never use your discussion to be disrespectful to other researchers or their work. This has been referred to as the “bully pulpit”. As with all aspects of professional interactions, consider how you would like to be treated, and act accordingly. This is not to say that you shouldn’t point out mistakes that were made before, but be sure not to become emotive or insulting.

@@

### Where next?
The ‘where next’ aspect of your discussion is important as it may provide the reader with ideas for their own work. Of course, these are questions that you may wish to pursue in your own career, or they may be corroborating evidence from other disciplines that you will never undertake. Either way, making pointers for continuing aspects of the research is an important component of the discussion. Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](results.html). This should be justifyable and not gratuitous. There's no point in suggesting what other studies could be done just for the sake of it, or just because you happen to have already done them in your thesis. But if there are legitate parts of your thesis that link to your discussion of your chapter (and this is quite likely), then you should definitely point to them.

### Last paragraph
The last paragraph of the discussion is your take home message. It’s a summary paragraph that sets out what you aimed to achieve, and what the new state of understanding of the topic is now that your [results](results.html) are out. This should include the key literature that can now be reconsidered.

#### Never repeat text
Please remember that while this might sound similar to your first paragraph, it is not the same. This final paragraph should not replicate any text that appears elsewhere in your chapter or manuscript (not even the abstract). Never repeat or copy text generally, even within your own chapter (or between chapters). For the reader, it’s very easy to spot and it gives the impression that you have nothing to say and are simply filling space. This is not the kind of impression that you want to give your reader, especially if they are examining your work!

As always, there are a number of other places to look for more advice to write your discussion, and I’d encourage you to read as widely as possible. For example, try Hess [-@hess2004how] and Sanli [-@sanli2013how].

>>>>

# The Discussion {#discussion}

@@

I'd like to repeat that the convention described above is a formula that is intended to help you getting started with writing. It is not the only way to write, and if you feel confident that another way is better then I encourage you to explore this with the support of your advisor. Be imaginative and don't feel that you are constrained or compelled to do anything in a formulaic way.

## First paragraph of the discussion
Your discussion begins by you responding to your hypothesis, clearly stating whether or not it has been accepted, and putting this into the wider context of the study (i.e. paragraph one or two of the [introduction](introduction.html) with relevant literature). You can then follow these statements by emphasising what you consider to be the most important finding, and explain how it adds to existing literature. However, don’t be tempted to over-interpret your results, or claim that they mean more than they do ([see section on speculation](speculation.html) below).

@@

## To sub-section or not to sub-section the discussion?
My preference is to plan the discussion before you write it, just as you did for the [introduction](introduction.html) using a bullet pointed list. This will provide you with logical sub-section headers for the discussion under which you can write the first draft. When your chapter has a simple aim that is easily communicated, I’d suggest deleting these sub-section headings before you finish. However, many studies are more complex and contain multiple experiments or evidential approaches. It is then sometimes wise to leave sub-sections in your discussion so that your reader can more easily follow the text. Where possible, these should be the same sub-sections that you have broken your [methods](methods.html) and [results](results.html) into, especially where these relate to specific hypotheses or aims. Or it may be more appropriate to discuss the different approaches separately, specifically when the literature that you refer to falls into different groups.

@@

In order to provide a critical evaluation of your research findings, you really need to present a reasoned argument [@jenicek2006how]. In some respects, when writing the discussion your walk a tightrope between inferring the meaning of your results for a wider audience, and making claims that are too broad (Table \@ref(tab:discuss)). Overreaching in the discussion (and consequently [abstract](abstract.html)) is a common problem that examiners and reviewers regularly ask authors pull back on their claims.

@@

Table: (\#tab:discuss) **Some discussion dos and don'ts (adapted from Jenicek 2006).** Balance your discussion so that you provide a reasoned argument that reflects what you found.

| **Do** | **Don't** |
|----|----|
| Provide a summary of your major findings | Use statements that are too broad |
| Interpret your results with respect to mechanisms | Over-represent your results |
| Give problems with methods and techniques | Provide unrealistic solutions to deficiencies |
| Compare similarities and contrasts with other studies | Ignore conflicting literature |
| Provide practical implications of your results | Be unrealistic about what could be done |
| Give directions for future research | Simply state what you have done elsewhere in your thesis |
| Restrain your conclusions to evidence from your data | Make unsubstantiated claims that overreach your evidence |
| Provide insight into unresolved questions, controversies and biases | [_Reductio ad absurdum_](argument.html)  |
| Explain the study design's limitations and uncertainties | Inflate the importance of your findings |
| Highlight the strengths of your study | Forget to mention the limitations |
| Provide alternative interpretations and hypotheses | Go into tangential issues |
| Give unexpected results and anomalies | Avoid being critical of how your study could have produced unexpected results |
| Mention literature that contradicts your findings | Use your conclusions as a 'bully pulpit' |
| Provide arguments that are logical and coherent | Erect a ['straw man' or 'false dichotomy'](argument.html) |
| Make conclusions based on your logic | Make emotional appeals to the reader |
| Make sure that you've discussed all results | Suddenly introduce new data or results not previously presented |
| Boast about your findings | Apologise or undermine your own work |
| Be concise | Waffle on |
| Take a look at target journal requirements and suggestions for a  discussion | Feel that journal advice is not relevant to you |

## Caveats and limitations
An important aspect of the discussion is to consider how the interpretation of your [results](results.html) may be incorrect. For example, if you have done an experiment, how well controlled was it and how well could it be considered to scale up to real world interactions? Could you have measured other variables? Almost every study will have caveats and limitations, and it is very important that you report them in a considered approach. You can also add ideas on how to improve your approach to eliminate important caveats that you have identified, even if this means suggesting something that adds evidence from a completely different field of biology.

@@

## Should you speculate in the discussion? {#speculation}
Reviewers will often be unhappy with speculation in the discussion section. Speculation isn’t that hard to spot, as it occurs when you make claims for which your [results](results.html) have no foundation. I think that it is healthy to have one or two statements that are speculative, but clearly labelled as such. After all, after writing this paper, you are going to be one of the world experts in the topic, and thus your deeper understanding is often worth relating to the reader. However, I suggest that you speculate in combination with suggesting what work could be done in future. If you really feel that the point has to be made, you must clearly label it as speculative.

@@

## Don’t beat up on others
Your [results](results.html) may show that other researchers were wrong with their interpretation or findings. Whatever you may think of them, never use your discussion to be disrespectful to other researchers or their work. This has been referred to as the “bully pulpit”. As with all aspects of professional interactions, consider how you would like to be treated, and act accordingly. This is not to say that you shouldn’t point out mistakes that were made before, but be sure not to become emotive or insulting.

Things brings to mind Freud's narcissism of minor differences, in which he pointed out that bitter fights could start between groups when they in fact agree on 90% on issues. So it is likely to be true that someone you might disagree with one one point, is likely to be allied with you on nearly all of their other views and opinions. Making and maintaining allegiances is more likely to get you in a position where you can constructively discuss any minor disagreements. Consider this before you are tempted to launch a negative sounding social media post.

@@

## Where next?
The ‘where next’ aspect of your discussion is important as it may provide the reader with ideas for their own work. Of course, these are questions that you may wish to pursue in your own career, or they may be corroborating evidence from other disciplines that you will never undertake. Either way, making pointers for continuing aspects of the research is an important component of the discussion. Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](results.html). This should be justifiable and not gratuitous. There's no point in suggesting what other studies could be done just for the sake of it, or just because you happen to have already done them in your thesis. But if there are legitimate parts of your thesis that link to your discussion of your chapter (and this is quite likely), then you should definitely point to them.

## Last paragraph
The last paragraph of the discussion is your take home message. It’s a summary paragraph that sets out what you aimed to achieve, and what the new state of understanding of the topic is now that your [results](results.html) are out. This should include the key literature that can now be reconsidered.

### Never repeat text
Please remember that while this might sound similar to your first paragraph, it is not the same. This final paragraph should not replicate any text that appears elsewhere in your chapter or manuscript (not even the abstract). Never repeat or copy text generally, even within your own chapter (or between chapters). For the reader, it’s very easy to spot and it gives the impression that you have nothing to say and are simply filling space. This is not the kind of impression that you want to give your reader, especially if they are examining your work!

As always, there are a number of other places to look for more advice to write your discussion, and I’d encourage you to read as widely as possible. For example, try Hess [-@hess2004how], Jenicek [-@jenicek2006how] and Şanlı [-@sanli2013how].

### 2021-04-19 3.08-References.Rmd
Add files via upload
# updated references with DOI and discussion table

## The References {#references}

@@

@@

### Digital Object Identifiers (DOIs) {#DOI}
A [DOI](https://www.doi.org/) is an [ISO](https://www.iso.org/) unique character string to identify physical, digital or abstract objects. Their beauty is that they are persistent over time, so once issued they will always be a way to reach a particular object.

@@

#### DOI Tools

@@

There's also an equally useful database that provides BibTex for DOIs that you enter: [doi2bib.org](https://doi2bib.org/)

>>>>

# The References {#references}

@@

Earlier in this book, I provided a guide to the differences between Vancouver and

@@

## Digital Object Identifiers (DOIs) {#DOI}
A [DOI](https://www.doi.org/) is an international standard ([ISO](https://www.iso.org/)) unique character string to identify physical, digital or abstract objects. Their beauty is that they are persistent over time, so once issued they will always be a way to reach a particular object.

@@

### DOI Tools

@@

There's an equally useful database that provides BibTex for DOIs that you enter: [doi2bib.org](https://doi2bib.org/)
